---
title: "Wine Quality Analysis"
subtitle: "data607- week 1 assingment"
date: "9/1/2026"
author: "DW"
format:
  html:
    toc: true
    codefold: true
  
editor: visual
---

# Wine Quality Analysis

## Approach

In this assignment I selected a datasat from uc irvine that contains information about different chemical attributes in wine. the data set contains 11 features and is a good way to try to handle numerical data for feature selection and classification or regression analysis.

My plan for this assignment is to load the data into R, and then do some slice some of the information into a smaller dataset of about 25%.

I can also do some simple EDA on the datasets

Data source:\
<https://raw.githubusercontent.com/DW8888/Wine-Quality/refs/heads/main/winequality-white.csv> <https://raw.githubusercontent.com/DW8888/Wine-Quality/refs/heads/main/winequality-red.csv>\
\

## Codebase

First we need to import our library and data sets:

```{r import-lib}
# Getting the tidyverse package
library("tidyverse")
```

```{r import-data}
# importing our data: my data set is seperated by the ; delimineter so i make sure to let the read dunction know to expect that.

red_wines<- read.csv("winequality-red.csv", sep=";")
white_wines<- read.csv("winequality-white.csv" , sep=";")
```

Lets see what our files gave us.\

```{r data-sum}
str(red_wines)
```

```{r data-sum2}
str(white_wines)
```

## Minipulation 

I want to set a label so i can then use this data set to predict wine color.

```{r data-mut}
# will make a factor colmn and attach the value 1 to red then combine the sets
red_wines$is_red <- factor(1, levels = c(0, 1), labels = c("white", "red"))

white_wines$is_red <- factor(0, levels = c(0, 1), labels = c("white", "red"))

wines <- rbind(red_wines, white_wines)
print(head(wines))
```

Summary of our data

```{r eda}
summary(wines)
```

```{r}
#| label: variable-distributions
#| fig-width: 12
#| fig-height: 16



numeric_vars <- names(wines)[sapply(wines, is.numeric)]

par(mfrow = c(5, 3), mar = c(3, 3, 2,1))
# itterate through the list and plot the variables

for (variable in numeric_vars) {
  hist(
    wines[[variable]],
    main = variable,
    xlab = "",
    col = "skyblue",
    border = "white"
  )
}




```

```{r}
barplot(
  table(wines$is_red),
  main = "Wine type",
  xlab = "is_red",
  ylab = "Count",
  col = c("lightyellow", "maroon"))
```
