---
title: "Wine Quality Analysis"
subtitle: "DATA 607: Week 1 Assignment Examplar"
date: "9/1/2026"
author: "DW"
format:
  html:
    toc: true
    codefold: true
  
editor: visual
---

# Wine Quality Analysis

## Approach

In this assignment, I selected a dataset from the UCI Machine Learning Repository that contains information about the chemical attributes of wine. The dataset contains 11 features and provides a useful opportunity to practice working with numerical data, feature selection, and classification or regression analysis.

My plan for this assignment is to load the data into R and create a label for the different wine types to use in a later classification task.

I will also perform some simple exploratory data analysis (EDA) on the datasets.

Data source:\
<https://raw.githubusercontent.com/DW8888/Wine-Quality/refs/heads/main/winequality-white.csv> <https://raw.githubusercontent.com/DW8888/Wine-Quality/refs/heads/main/winequality-red.csv>\
\
Cortez, P., Cerdeira, A., Almeida, F., Matos, T., & Reis, J. (2009). Wine Quality \[Dataset\]. UCI Machine Learning Repository. <https://doi.org/10.24432/C56S3T.>

## Research Question

Can the chemical attributes of a wine be used to classify it as red or white?

## Codebase

First, we need to import the library and datasets:

```{r import-lib}
# Getting the tidyverse package
library("tidyverse")
```

```{r import-data}
# Importing the data: the dataset uses semicolons as delimiters, so I specify that in the read function.

red_wines<- read.csv("winequality-red.csv", sep=";")
white_wines<- read.csv("winequality-white.csv" , sep=";")
```

Let's inspect the structure of the datasets.\

```{r data-sum}
str(red_wines)
```

```{r data-sum2}
str(white_wines)
```

## Manipulation

I want to create a label that I can use to predict wine color.

```{r data-mut}
# Create a factor column, assign 1 to red wine, and then combine the datasets.
red_wines$is_red <- factor(1, levels = c(0, 1), labels = c("white", "red"))

white_wines$is_red <- factor(0, levels = c(0, 1), labels = c("white", "red"))

wines <- rbind(red_wines, white_wines)
print(head(wines))
```

Here is a summary of the data:

```{r eda}
summary(wines)
```

```{r}
#| label: variable-distributions
#| fig-width: 12
#| fig-height: 16



numeric_vars <- names(wines)[sapply(wines, is.numeric)]

par(mfrow = c(5, 3), mar = c(3, 3, 2,1))
# Iterate through the list and plot the variables.

for (variable in numeric_vars) {
  hist(
    wines[[variable]],
    main = variable,
    xlab = "",
    col = "skyblue",
    border = "white"
  )
}




```

```{r}
barplot(
  table(wines$is_red),
  main = "Wine type",
  xlab = "is_red",
  ylab = "Count",
  col = c("lightyellow", "maroon"))
```

## Tidyverse Comparison

The following summary compares the number of wines, average alcohol content, and average quality for each wine type.

```{r wine-type-summary}
wines |>
  group_by(is_red) |>
  summarise(
    number_of_wines = n(),
    average_alcohol = round(mean(alcohol), 2),
    average_quality = round(mean(quality), 2)
  )
```

```{r alcohol-by-wine-type}
ggplot(wines, aes(x = is_red, y = alcohol, fill = is_red)) +
  geom_boxplot() +
  scale_fill_manual(values = c('white' = 'lightyellow', 'red' = 'maroon')) +
  labs(
    title = "Alcohol Content by Wine Type",
    x = "Wine type",
    y = "Alcohol content"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Thoughts

\
The dataset has no missing values and contains 11 numerical variables for each wine. I added a new variable to indicate whether the wine is red or white. Later, I plan to build a classification model that predicts whether a wine is red or white based on its features. It is important to note that the dataset has a class imbalance between red and white wines, which I will need to account for.\
Many of the variables do not follow a normal distribution and would have to be normalized if we would like to do some sort of linear regression modeling.